---
title: "EDS 222: Homework 2"
format: 
  html:
    toc: true
---

```{r}
#| label: setup
#| include: false

# Load required packages
library(tidyverse)

```

## Question 1: Probability density functions in `R`

`R` has many built-in functions that let you describe, analyze, and sample from common probability density functions. For example, if you type `?stats::Normal` in your console, you'll see documentation on all the functions relevant to the normal distribution. These functions include:

-   `dnorm(x, mean = 0, sd = 1)`, which returns the *density* of the normal distribution evaluated at whatever values you pass in via `x`. You can think of the output of `dnorm()` as the *height* of the normal pdf at the value `x`. Note that this function defaults to a normal distribution with $\mu = 0$ and $\sigma = 1$, but you can of course change that.

-   `pnorm(q, mean = 0, sd = 1)`, which returns the *cumulative probability* of the normal distribution evaluated at whatever values you pass in via `q`. You can think of the output of `pnorm()` as the *area* under the pdf to the left of the value `q`. Again, note the default distribution parameters $\mu$ and $\sigma$.

::: callout-note
Recall that the normal distribution is a family of distributions that are symmetric and do not have long tails. They each have different means $\mu$ and standard deviations $\sigma$.
:::

### Question 1.1

Use `dnorm()` to compute the density of the normal pdf for all values in the `x` vector generated below, using $\mu = 0$ and $\sigma = 1$. Use `geom_polygon()`, `geom_line()`, or `geom_point()` (take your pick) to plot this pdf over the support given in `x`.

```{r}
#| label: q1-1

x <- seq(-4, 4, 0.01)

```

### Question 1.2

Use the densities you generated in **Question 1.1** to calculate the probability that a random variable distributed normally with mean 0 and standard deviation 1 falls between -2 and 2.

::: callout-hint
Remember that $$ Pr(A\leq x \leq B) = \int_A^B f(x)dx $$where the integral is a fancy way to tell you to sum up $f(x)$ over values of $x$ from $A$ to $B$.
:::

```{r}
#| label: q1-2



```

### Question 1.3

Suppose $\sigma=2$ instead. Qualitatively, how would your answer to Question 1.2 change? Why?

### Question 1.4

An analogous set of functions computes densities and probabilities for the **log normal** distribution. These functions are `dlnorm()` and `plnorm()` and operate as above for the normal distribution functions.

Use `plnorm()` under default parameters to compute the probability that a random variable distributed log normal takes on a value above 2. Use `pnorm()` to compute the corresponding probability for the normal distribution under default parameters. Why are these values so different?

```{r}
#| label: q1-3



```

## Question 2: Climate summary statistics

In the following questions, you'll be working with climate data from Colombia. These data were obtained from the [ERA5 database](https://www.ecmwf.int/en/forecasts/datasets/reanalysis-datasets/era5), a product made available by the European Centre for Medium-Range Weather Forecast. The high-resolution hourly gridded data were aggregated to the municipality by month level -- that is, each observation in these data report a monthly average temperature value and a monthly cumulative precipitation value for one of the 1,123 municipalities across the country.

::: callout-note
The computational techniques we use to go from raw, spatial, gridded data to a tabular dataset at an administrative level are really valuable for environmental data science. You will learn more about these topics in EDS 223!
:::

These data -- stored in `colombia_climate.csv` -- cover all municipalities for the period 1996 to 2015. Climate scientists tend to describe the "climate" of a location as the probability density function of a large set of climate variables over about a 30 year period. We only have 20 years, but we will consider our sample as randomly drawn temperature and precipitation realizations from the "climate" p.d.f. over this period. We are aiming to draw conclusions about the Colombian climate using this sample of temperature and precipitation observations.

### Question 2.1

Read these data into `R` using the `read_csv()` function.

::: callout-tip
See the README for details on the variables in `colombia_climate.csv`.
:::

For each of the temperature and rainfall variables, create a histogram that shows the distribution of the variable across the entire sample. For each variable, answer the following questions:

-   Is the distribution symmetric or skewed?
-   Is there a long tail (or two), or does this distribution look approximately normally distributed?
-   Is the distribution unimodal, bimodal, or multimodal?

```{r}
#| label: q2-2



```

### Question 2.2

Given your answers to 2.1 above, do you expect the mean of temperature to differ from the median? Is it likely to be about the same, smaller, or larger? What about precipitation?

### Question 2.3

Anthropogenic climate change is expected to raise temperatures across Colombia, increase total precipitation, and increase variability in precipitation. Compute the mean, the median, and the standard deviation of each climate variable in:

-   All years before and including 2005
-   All years after 2005

Put your summary statistics into a table (or two tables, whatever is easiest). Are the changes you see between the pre-2005 and post-2005 periods consistent with climate change? Explain why.

### Question 2.4

The histograms and summary statistics should make you concerned that these data are not normally distributed. As we will show later in the course, it's often very helpful to have normally distributed data before we do things like linear regressions or hypothesis testing. Here, let's use a Q-Q plot to assess the normality of our sample data.

-   Use `geom_qq()` and `geom_qq_line()` in `ggplot2` to make a Q-Q plot for each variable.

-   What do you conclude about the normality of these two variables?

::: callout-tip
`geom_qq_line()` lets you draw a line indicating where the sample quantiles would lie if the data were normally distributed.
:::

```{r}
#| label: q2-4



```

### Question 2.5

When our sample observations are not normally distributed, we often rely on nonlinear transformations to reshape our data. If we compute a nonlinear transformation on our underlying data and they then look closer to normal, we can use this transformed version of our variable in later statistical analysis.

::: callout-note
Any mathematical operation that is a nonlinear function of the underlying variable can be considered a "nonlinear transformation". For example, $x^2$ and $log(x)$ are both nonlinear transformations.
:::

Because we tend to see a lot of variables in the world that follow the lognormal distribution, a very common nonlinear transformation is the natural logarithm. Transform the precipitation data by taking the natural logarithm. Then remake your Q-Q plot -- does your variable (defined as `log(precip)`) now look closer to normally distributed? What can you learn about where the data diverge from the normal distribution?
